\jaxcosmo: An End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2023-02-05 paper_OJA.tex
Update the paper

%%%% new command (JEC)
\newcommand{\nn}{\nonumber}
\newcommand{\numpyro}{\texttt{NumPyro}}
\newcommand{\bydef}{:=}
\newcommand{\jaxcosmo}{\texttt{jax-cosmo}}
\newcommand{\autodiff}{\texttt{autodiff}}

@@

\author{
J.~E. Campagne$^{1,\ast}$, 
F. Lanusse$^2$,
J. Zuntz$^3$,\\
A. Boucaud$^8$,
D.~Kirkby$^4$,
D. Lanzieri$^2$,
Y.~Li$^{5,6}$,
A. Peel$^7$
\jaxcosmo\ contributors}
\thanks{$^\ast$jean-eric.campagne@ijclab.in2p3.fr}

\affiliation{
$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France\\
}

\affiliation{$^5$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
\affiliation{$^6$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
\affiliation{$^7$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}
\affiliation{$^8$Université de Paris, CNRS, Astroparticule et Cosmologie, F-75013 Paris, France}

@@

\begin{abstract}

We present \jaxcosmo, a library for differentiable cosmological theory calculations. \jaxcosmo\ uses the recent \texttt{JAX} library, which has created a new coding ecosystem, especially in probabilistic programming.
%
As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \texttt{JAX} exposes an \textit{automatic differentiation} (AD) mechanism. AD can take a Python function that uses \texttt{JAX} and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn. This gives us the opportunity to apply a range of powerful Bayesian inference algorithms, otherwise impractical in cosmology, such as Hamiltonian Monte Carlo and Variational Inference. 
%
\jaxcosmo\ implements background evolution, linear and non-linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes, all differentiable with respect to the cosmological parameters and their other inputs. 
%
We discuss algorithms made possible by this library, and present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference. 
%
We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that Normalizing Flows using the Neural Transport are a promising methodology.

% The recent \texttt{JAX} library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems. In particular, \texttt{JAX} exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses \texttt{JAX} and return its (vector) derivative.  This gives us the opportunity to apply a range of powerful but otherwise unfeasible algorithms used in Bayesian inference, such as Hamiltonian Monte Carlo (HMC) and Variational Inference.

@@

\section{Introduction}
Bayesian inference has been widely used in cosmology in the form of Monte Carlo Markov Chains (MCMC) since the work of \citep{2001ApJ...563L..95K,2003MNRAS.341.1084R} and has been the keystone for past and current analysis thanks partly to packages as  \texttt{CosmoMC} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS} \citep{2015A&C....12...45Z}, \texttt{MontePython} \citep{2019PDU....24..260B} as well as   \texttt{Cobaya} \citep{2019ascl.soft10019T,2021JCAP...05..057T}; see, for instance, the list of citations to these popular packages for an idea of the wide usage in the community.

@@

While these tools have especially been applied to neural network optimization and machine learning (ML) in general, they can also enable classical statistical methods that require the derivatives of (e.g. likelihood) functions to operate: we consider such methods in this paper. \textit{Autodiff} has been implemented in widely used libraries like \texttt{Stan} \citep{JSSv076i01}, \texttt{TensorFlow}  \citep{tensorflow2015-whitepaper}, and  \texttt{PyTorch} \citep{NEURIPS2019_9015}.

A recent entrant to this field is the \texttt{JAX} library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes. \texttt{JAX} offers an easy parallelization mechanism (\texttt{vmap}), and just-in-time compilation (\texttt{jit}) and optimization targetting CPU, GPU and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow. These attractive features have driven wide adoption of \texttt{JAX} in computational research, and motivate us to consider its usage in cosmology.

\texttt{JAX} contains bespoke reimplementation of packages such as \texttt{jax.numpy} and \texttt{jax.scipy}, as well as example libraries as \texttt{Stax} for simple but flexible neural network development. It has formed the seed for a wider ecosystem of packages, including, for example: 
\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} an hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro\ \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper. Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to \texttt{JAX} backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.

@@

In this context we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper. The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable \texttt{JAX} functions.

@@

\section{JAX: GPU Accelerated and Automatically Differentiable Python Programming}
\label{sec-primer}

The aim of this section is to provide a brief technical primer on \texttt{JAX}, necessary to fully grasp the potential of a cosmology library implemented in this framework.

@@

\paragraph{\textbf{Automatic Differentiation}} Traditionally, two different approaches have been used in cosmology to obtain derivatives of given numerical expressions. The first approach is to derive analytically the formula for the derivatives of interest \citep[e.g.][]{2013MNRAS.432..894J}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}. This is however only practical typically for simple analytical models.
The second approach is to compute numerical derivatives by finite differences. This approach can be applied on black-box models of arbitrary complexity (from typical Boltzmann codes to cosmological simulations \citep{2020ApJS..250....2V}). However it is notoriously difficult to obtain stable derivatives by finite differences \citep[e.g.][]{2021arXiv210100298B, 2021A&A...649A..52Y}. In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them impractical whenever derivatives are needed as part of outer iterative algorithm.

Automatic differentiation frameworks like \texttt{JAX} take a different approach. They trace the execution of a given model and decompose this trace into fundamental operations with known derivatives (e.g. the derivative of a multiplication operation is known). Then, by applying the chain rule formula, the computational graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operations. A new function corresponding to the derivative of the original function is therefore built automatically for the user. We direct the interested reader to \cite{baydin2018automatic, Margossian2019} for in-depth introductions to automatic differentiation.

JAX provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in \texttt{JAX}.
\begin{lstlisting}[language=iPython]
# Let us define a simple function 
def f(x):
	return y = 5 * x + 2
# Take the derivative
df_dx = jax.grad(f)
# df_dx is a new function that always return 5 
\end{lstlisting}

\textbf{Why is that interesting?} \autodiff\ makes it possible to obtain \textit{exact gradients of cosmological likelihoods} with respect to all input parameters at the cost of only 2 likelihood evaluations.

\paragraph{\textbf{Just In Time Compilation (JIT)}} Despite its convenience and wide adoption in astrophysics, Python still suffers from slow execution times compared to fully compiled languages such as C/C++. One approach to mitigate these issues and make Python code fast is Just In Time compilation, which traces the execution of a given Python function the first time it is called, and compiles it into a fast executable (by-passing the Python interpreter), which can be transparently used in subsequent calls to this function. Compared to other strategies for speeding up Python code such as Cython, JIT allows the user to simply write plain Python code, and reap the benefits of compiled code.

A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba\footnote{\url{https://numba.pydata.org/}}, or the HOPE library \cite{2015A&C....10....1A} developed specifically for the need of astrophysics. \texttt{JAX} stands out compared to these other frameworks in that it relies on the highly optimized XLA library\footnote{\url{https://www.tensorflow.org/xla}} for executing the compiled expressions. XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU an Tensor Processing Units (TPU) clusters. The ability to perform computations directly on GPUs through XLA is one of the major benefits of \texttt{JAX}, as speed-ups of at least two orders of magnitudes can be expected for typical parallel linear algebra computations compared to CPU.

In \texttt{JAX}, jitting is achieved simply by transforming a function with \texttt{jax.jit} operation:
\begin{lstlisting}[language=iPython]
# Let us redefine our function
def f(x):
	return y = 5 * x + 2
# And JIT it
jitted_f = jax.jit(f)
# The first call to jitted_f will be relatively slow
# subsequent calls will be extremely fast
# and run as a compiled code directly on GPU
\end{lstlisting}

\textbf{Why is that interesting?} JIT makes it possible to execute entire cosmological \textit{MCMC chains directly on GPUs as compiled code}, with orders of magnitude in speedup over Python code.

\paragraph{\textbf{Automatic Vectorization}} Another extremely powerful feature of \texttt{JAX} is its ability to automatically vectorize or paralellize any function. Through the same tracing mechanism used for automatic differentiation, \texttt{JAX} can decompose a given computation into fundamental operations and add a new \textit{batch} dimension so that the computation can be applied on a batch of inputs as opposed to individual inputs. In doing so, note that the computation will not run sequentially over all entries of the batch, but trully in parallel making full use of the intrinsic parallel architecture of modern GPUs.

In \texttt{JAX} automatic vectorization is achieved using the \textit{jax.vmap} operation:
\begin{lstlisting}[language=iPython]
# Our function f only applies to scalars
def f(x):
	return y = 5 * x + 2
# Applying automatic vectorization
batched_f = jax.vmap(f)
# batched_f now applies to 1D arrays
\end{lstlisting}
Again, we stress that in this example, \texttt{batched\_f} will not be implemented in terms of a for loop, but with operations over vectors. The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood. When considering multi-devices use-case (eg. GPU or TPU), \texttt{JAX} provides \texttt{pmap} which compile and execute in parallel replicates of the same code on each device. Moreover, recent experimental developments deal with parallelization of function over supercomputer-scale hardware meshes. Notice that in the examples detailed in this article, we have only relied on \texttt{vmap} functionality.

\textbf{Why is that interesting?} Automatic Vectorization makes it possible to trivially parallelize cosmological likelihood evaluations, to run many parallel MCMC chains on a single GPU.

\paragraph{\textbf{NumPy API compliance}} Finally, the last point to note about \texttt{JAX}, is that it mirrors the NumPy API (with only a few exceptions). This means in practice that existing NumPy code can easily be converted to \texttt{JAX}. This is in contrast to other similar frameworks like TensorFlow, PyTorch, or Julia which all require the user to learn, and adapt their code to, a new API or even a new language.

\textbf{Why is that interesting?} NumPy compliance implies improved maintainability and lower barrier to entry for new contributors.

\section{Design of the \jaxcosmo\ library}
\label{sec-jaxcosmo-design}

In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in JAX. The general design follows that of the Core Cosmology Library (CCL) \citep{2019ApJS..242....2C}, though, in its initial release,
 \jaxcosmo\ only implements a subset of CCL features and options.

All \jaxcosmo\ data structures are organized as \texttt{JAX} \textit{pyTree} container objects, ensuring that two key \texttt{JAX} features are available to them: \textit{autodiff} and \textit{vmap}. The vmap feature enables any operation defined in \texttt{JAX} (including complicated composite operations) to be applied efficiently as a vector operation. The autodiff feature further makes it possible to take the derivative of any operation, automatically transforming a function that takes $n$ inputs and produces $m$ outputs into a new function that generates an $m \times n$ matrix of partial derivatives.

\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, and Simpson \& Romberg integration routines through this framework, so that we can automatically compute the derivatives of their solutions with respect to their inputs. This includes not only the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also other input quantities such as redshift distributions as described below.

In the rest of this section we describe the cosmological calculations that are implemented using these facilities.

\subsection{Background cosmology}

The computation of the evolution of the cosmological background follows a typical implementation of a Friedmann equation (\citealp[see e.g.][]{2005A&A...443..819P}):
\begin{equation}
    E^2(a) = \frac{\mathrm{H}^2(a)}{\mathrm{H}^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
\end{equation}
with $a=1/(1+z)$ the scale factor related to the redshift $z$, $\mathrm{H}(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and 
\begin{equation}
    f(a) = -3 (1 + w_0 + w_a) \ln(a) + 3 w_a (a - 1)
\end{equation}
Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are currently neglected. From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions, such as the radial comoving distance
\begin{equation}
     \chi(a) =  R_H \int_a^1 \frac{da^\prime}{{a^\prime}^2 E(a^\prime)}
     \label{eq:radial_comoving}
\end{equation}
with $R_H$ the Hubble radius. 
%Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
%
\subsection{Growth of perturbations}
%
Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function $T$ which transforms the primordial matter power spectrum to its late-time non-linear value:
\begin{equation}
    P(k, z) = P(k, z=\infty) \cdot T(k, z; \Omega_m, \Omega_b, ...),
\end{equation}
through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}. No Baryon feedback modeling is considered yet.

@@

The normalisation $A$ is parametrized via $\sigma_8$ at $z=0$ as 
\begin{equation}
    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{dk}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
\end{equation}
with $R_8 = 8 \mathrm{Mpc}/h$ and $W(x)$ related to the $j_1$ spherical Bessel function as
\begin{equation}
    W(x) = \frac{3j_1(x)}{x}
\end{equation}

%Future version of the library would offer the possibility to call a \texttt{JAX} emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}. \JZ{How close is this to being usable? If it's just an idea or an early prototype then skip this paragraph.}
%
\subsection{Angular power spectra}
%
\jaxcosmo\ is currently focused on predicting projected 2D Fourier-space 2pt galaxy lensing, clustering, and cross correlations, the $C_\ell$ angular power spectra that are a primary target of upcoming photometric surveys. The details of the implementation is in \texttt{jc.angular\_cl} which deals with the mean and Gaussian covariance matrix computations.

>>>>

%%%% new command (JEC)
\newcommand{\nn}{\nonumber}
\newcommand{\numpyro}{\texttt{NumPyro}}
\newcommand{\bydef}{:=}
\newcommand{\jaxcosmo}{\texttt{jax-cosmo}}
\newcommand{\autodiff}{\texttt{autodiff}}
\newcommand{\jax}{\texttt{JAX}}

@@

\author{
J.~E. Campagne$^{1,\ast}$, 
F. Lanusse$^2$,
J. Zuntz$^3$,\\
A. Boucaud$^4$,
S. Casas$^{5}$,
M.~Karamanis$^{6,7}$,
D.~Kirkby$^8$,
D. Lanzieri$^9$,
Y.~Li$^{10,11}$,
A. Peel$^{12}$
}
\thanks{$^\ast$jean-eric.campagne@ijclab.in2p3.fr}

\affiliation{
$^1$Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
}
\affiliation{$^2$Université Paris-Saclay, Université Paris Cité, CEA, CNRS, AIM, 91191, Gif-sur-Yvette, France}
\affiliation{$^3$Institute for Astronomy, University of Edinburgh, Edinburgh EH9 3HJ, United Kingdom}
\affiliation{$^4$Université de Paris, CNRS, Astroparticule et Cosmologie, F-75013 Paris, France}
\affiliation{$^{5}$Institute for Theoretical Particle Physics and Cosmology (TTK), RWTH Aachen University, 52056 Aachen, Germany.}
\affiliation{$^{6}$Berkeley Center for Cosmological Physics, University of California, Berkeley, CA 94720, USA}
\affiliation{$^{7}$Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley, CA 94720, USA}
\affiliation{$^8$Department of Physics and Astronomy, University of California, Irvine, CA 92697, USA}
\affiliation{$^9$Université Paris Cité, Université Paris-Saclay, CEA, CNRS, AIM, F-91191, Gif-sur-Yvette, France}
\affiliation{$^{10}$Department of Mathematics and Theory, Peng Cheng Laboratory, Shenzhen, Guangdong 518066, China}
\affiliation{$^{11}$Center for Computational Astrophysics \& Center for Computational Mathematics, Flatiron Institute, New York, New York 10010, USA}
\affiliation{$^{12}$Ecole Polytechnique F\'ed\'erale de Lausanne (EPFL), Observatoire de Sauverny, 1290 Versoix, Switzerland}

@@

\begin{abstract}
We present \jaxcosmo, a library for automatically differentiable cosmological theory calculations. \jaxcosmo\ uses the \jax\ library, which has created a new coding ecosystem, especially in probabilistic programming.
%
As well as batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU), \jax\ exposes an \textit{automatic differentiation} (AD) mechanism. Thanks to AD, \jaxcosmo\ gives access to the derivatives of cosmological likelihoods with respect to any of their parameters, and thus enables a range of powerful %  AD can take a Python function that uses \jax\ and return its (vector) derivative, computed not with finite differences but by successively differentiating each instruction in turn. This gives us the opportunity to apply a range of powerful
Bayesian inference algorithms, otherwise impractical in cosmology, such as Hamiltonian Monte Carlo and Variational Inference. 
%
In its initial release, \jaxcosmo\ implements background evolution, linear and non-linear power spectra (using \texttt{halofit} or the Eisenstein and Hu transfer function), as well as angular power spectra ($C_\ell$) with the Limber approximation for galaxy and weak lensing probes, all differentiable with respect to the cosmological parameters and their other inputs. 
%
We illustrate how automatic differentiation can be a game-changer for common tasks involving Fisher matrix computations, or full posterior inference with gradient-based techniques (e.g. Hamiltonian Monte Carlo). In particular, we show how Fisher matrices are now fast, exact, no longer require any fine tuning, and are themselves differentiable with respect to parameters of the likelihood enabling complex survey optimization by simple gradient descent. Finally, using a Dark Energy Survey Year 1 3x2pt analysis as a benchmark, we demonstrate how \jaxcosmo\ can be combined with Probabilistic Programming Languages such as \numpyro\ to perform posterior inference with state-of-the-art algorithms including a No U-Turn Sampler (NUTS), Automatic Differentiation Variational Inference (ADVI), and Neural Transport HMC (NeuTra).% We discuss algorithms made possible by this library, present comparisons with the Core Cosmology Library as a benchmark, and run a series of tests using the Dark Energy Survey Year 1 3x2pt analysis with the \numpyro\ library to demonstrate practical inference. 
%
We show that clear improvements are possible using HMC compared to Metropolis-Hasting, and that Normalizing Flows using the Neural Transport are a promising methodology.\FrL{we need a better statement than this, what are our conclusions? Are we faster/better than Cobaya?}

% The recent \jax\ library has created a new ecosystem from which probabilistic programming software can take enormous benefit: batch acceleration, just-in-time compilation, and automatic optimization of code for different hardware modalities (CPU, GPU, TPU) can provide huge speed-ups to a wide range of different problems. In particular, \jax\ exposes an \textit{automatic differentiation} mechanism, which can take a python function that uses \jax\ and return its (vector) derivative.  This gives us the opportunity to apply a range of powerful but otherwise unfeasible algorithms used in Bayesian inference, such as Hamiltonian Monte Carlo (HMC) and Variational Inference.

@@

\section{Introduction}
Bayesian inference has been widely used in cosmology in the form of Monte Carlo Markov Chains (MCMC) since the work of \citet{2001ApJ...563L..95K} and \citet{2003MNRAS.341.1084R}, and has been the keystone for past and current analysis thanks partly to packages such as  \texttt{CosmoMC} \citep{2002PhRvD..66j3511L}, \texttt{CosmoSIS} \citep{2015A&C....12...45Z}, \texttt{MontePython} \citep{2019PDU....24..260B}, and \texttt{Cobaya} \citep{2019ascl.soft10019T,2021JCAP...05..057T}; see, for instance, the list of citations to these popular packages for an idea of the wide usage in the community.

@@

While these tools have especially been applied to neural network optimization and machine learning (ML) in general, they can also enable classical statistical methods that require the derivatives of (e.g. likelihood) functions to operate: we consider such methods in this paper. \textit{Autodiff} has been implemented in widely used libraries like \texttt{Stan} \citep{JSSv076i01}, \texttt{TensorFlow} \citep{tensorflow2015-whitepaper}, \texttt{Julia} \citep{bezanson2017julia}, and \texttt{PyTorch} \citep{NEURIPS2019_9015}.

A recent entrant to this field is the \jax\ library\footnote{\url{https://jax.readthedocs.io}} \citep{jax2018github} which has undergone rapid development and can automatically differentiate native \texttt{Python} and \texttt{NumPy} functions, offering a speed up to the development process and indeed code runtimes. \jax\ offers an easy parallelization mechanism (\texttt{vmap}), just-in-time compilation (\texttt{jit}), and optimization targeting CPU, GPU, and TPU hardware thanks to the \texttt{XLA} library that powers TensorFlow. These attractive features have driven wide adoption of \jax\ in computational research, and motivate us to consider its usage in cosmology.

\jax\ contains bespoke reimplementations of packages such as \texttt{jax.numpy} and \texttt{jax.scipy}, as well as example libraries such as \texttt{Stax} for simple but flexible neural network development. It has formed the seed for a wider ecosystem of packages, including, for example: 
\texttt{Flax} \citep{flax2020github} a high-performance neural network library, \texttt{JAXopt} \citep{jaxopt_implicit_diff} a hardware accelerated, batchable and differentiable collection of optimizers, \texttt{Optax} \citep{optax2020github} a gradient processing and optimization library, and \numpyro\ \citep{phan2019composable,bingham2019pyro}, a probabilistic programming language (PPL) that is used in this paper. Other PPL packages such as \texttt{PyMC} \citep{Salvatier2016} have switched to a \jax\ backend in recent versions\footnote{A more exhaustive list and rapidly growing list of packages can be found at \url{https://project-awesome.org/n2cholas/awesome-jax}}.

@@

In this context we have developed the open source \jaxcosmo\ library\footnote{\url{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}}, which we present in this paper. The package represents a first step in making the powerful features described above useful for cosmology; it implements a selection of theory predictions for key cosmology observables as differentiable \jax\ functions.

@@

\section{\jax: GPU Accelerated and Automatically Differentiable Python Programming}
\label{sec-primer}

The aim of this section is to provide a brief technical primer on \jax, necessary to fully grasp the potential of a cosmology library implemented in this framework.

@@

\paragraph{\textbf{Automatic Differentiation}} Traditionally, two different approaches have been used in cosmology to obtain derivatives of given numerical expressions. The first approach is to derive analytically the formula for the derivatives of interest \citep[e.g.][]{2013MNRAS.432..894J}, with or without the help of tools such as \texttt{Mathematica}\footnote{\url{https://www.wolfram.com/mathematica}.}. This is typically only practical, however, for simple analytical models.
The second approach is to compute numerical derivatives by finite differences. This approach can be applied on black-box models of arbitrary complexity (from typical Boltzmann codes to cosmological simulations; \citealp{2020ApJS..250....2V}). However it is notoriously difficult to obtain stable derivatives by finite differences \citep[e.g.][]{2021arXiv210100298B, 2021A&A...649A..52Y}. In addition, their computational cost does not scale well with the number of parameters (a minimum of $(2N+1)$ model evaluations is typically required for $N$ parameters), making them impractical whenever derivatives are needed as part of an outer iterative algorithm.

Automatic differentiation frameworks like \jax\ take a different approach. They trace the execution of a given model and decompose this trace into primitive operations with known derivatives (e.g. multiplication). Then, by applying the chain rule formula, the computational graph for the derivatives (of any order) of the model can be built from the known derivatives of every elementary operation. A new function corresponding to the derivative of the original function is therefore built automatically for the user. We direct the interested reader to \citet{baydin2018automatic} and \citet{Margossian2019} for in-depth introductions to automatic differentiation.

\jax\ provides in particular a number of operators (\texttt{jax.grad}, \texttt{jax.jacobian}, \texttt{jax.hessian}) which can compute derivatives of any function written in \jax:
\begin{lstlisting}[language=iPython]
# Define a simple function 
def f(x):
	return y = 5 * x + 2
# Take the derivative
df_dx = jax.grad(f)
# df_dx is a new function that always returns 5 
\end{lstlisting}

\textbf{Why is this interesting?} \autodiff\ makes it possible to obtain \textit{exact gradients of cosmological likelihoods} with respect to all input parameters at the cost of only two likelihood evaluations.

\paragraph{\textbf{Just In Time Compilation (JIT)}} Despite its convenience and wide adoption in astrophysics, Python still suffers from slow execution times compared to fully compiled languages such as C/C++. One approach to mitigate these issues and make Python code fast is Just In Time compilation, which traces the execution of a given Python function the first time it is called, and compiles it into a fast executable (by-passing the Python interpreter), which can be transparently used in subsequent calls to this function. Compared to other strategies for speeding up Python code such as Cython, JIT allows the user to write plain Python code, and reap the benefits of compiled code.

A number of libraries allowing for JIT have already been used in astrophysics, in particular Numba\footnote{\url{https://numba.pydata.org/}}, or the HOPE library \cite{2015A&C....10....1A} developed specifically for the needs of astrophysics. \jax\ stands out compared to these other frameworks in that it relies on the highly optimized XLA library\footnote{\url{https://www.tensorflow.org/xla}} for executing the compiled expressions. XLA is continuously developed by Google as part of the TensorFlow project, for efficient training and inference of large scale deep learning applications, and as such supports computational backends such as GPU and Tensor Processing Units (TPU) clusters. The ability to perform computations directly on GPUs through XLA is one of the major benefits of \jax, as speed-ups of at least two orders of magnitudes can be expected for typical parallel linear algebra computations compared to CPU.

In \jax, jitting is achieved by transforming a function with the \texttt{jax.jit} operation:
\begin{lstlisting}[language=iPython]
# Redefine our function
def f(x):
	return y = 5 * x + 2
# And JIT it
jitted_f = jax.jit(f)
# The first call to jitted_f will be relatively slow
# subsequent calls will be extremely fast
# and run as a compiled code directly on GPU
\end{lstlisting}

\textbf{Why is this interesting?} JIT makes it possible to execute entire cosmological \textit{MCMC chains directly on GPUs as compiled code}, with orders of magnitude gain in speedup over Python code.

\paragraph{\textbf{Automatic Vectorization}} Another extremely powerful feature of \jax\ is its ability to automatically vectorize or paralellize any function. Through the same tracing mechanism used for automatic differentiation, \jax\ can decompose a given computation into primitive operations and add a new \textit{batch} dimension so that the computation can be applied to a batch of inputs as opposed to individual ones. In doing so, the computation will not run sequentially over all entries of the batch, but truly in parallel making full use of the intrinsic parallel architecture of modern GPUs.

In \jax\ automatic vectorization is achieved using the \textit{jax.vmap} operation:
\begin{lstlisting}[language=iPython]
# Our function f only applies to scalars
def f(x):
	return y = 5 * x + 2
# Applying automatic vectorization
batched_f = jax.vmap(f)
# batched_f now applies to 1D arrays
\end{lstlisting}
Again, we stress that in this example, \texttt{batched\_f} will not be implemented in terms of a for loop, but with operations over vectors. The function above is trivial, but the same approach can be used to parallelize any function, from Limber integrals, to an entire likelihood. For multi-device use-cases (e.g., several GPUs or TPUs), \jax\ provides \texttt{pmap} which compiles and executes, in parallel, replicas of the same code on each device. Moreover, recent experimental developments deal with parallelization of functions over supercomputer-scale hardware meshes. In the examples detailed in this article, we have only relied on \texttt{vmap} functionality.

\textbf{Why is this interesting?} Automatic Vectorization makes it possible to trivially parallelize cosmological likelihood evaluations, to run many parallel MCMC chains on a single GPU.

\paragraph{\textbf{NumPy API compliance}} Finally, the last point to note about \jax, is that it mirrors the NumPy API (with only a few exceptions). This means in practice that existing NumPy code can easily be converted to \jax. This is in contrast to other similar frameworks like TensorFlow, PyTorch, or Julia which all require the user to learn, and adapt their code to, a new API or even a new language.

\textbf{Why is this interesting?} NumPy compliance implies improved maintainability and lower barrier to entry for new contributors.

\section{Capabilities of the \jaxcosmo\ library}
\label{sec-jaxcosmo-design}

In this section, we describe the cosmological modeling provided by \jaxcosmo, and its implementation in \jax. The general design follows that of the Core Cosmology Library (CCL; \citealp{2019ApJS..242....2C}), though in its initial release
 \jaxcosmo\ only implements a subset of CCL features and options.

All \jaxcosmo\ data structures are organized as \jax\ container objects, which means that two key \jax\ features are automatically available to them: \textit{autodiff} and \textit{vmap}. The vmap feature enables any operation defined in \jax\ (including complicated composite operations) to be applied efficiently as a vector operation. The autodiff feature further makes it possible to take the derivative of any operation, automatically transforming a function that takes $n$ inputs and produces $m$ outputs into a new function that generates an $m \times n$ matrix of partial derivatives.

\jaxcosmo\ implements, for example, Runge-Kutta solvers for ODEs, as well as Simpson and Romberg integration routines through this framework, so that we can automatically compute the derivatives of their solutions with respect to their inputs. This includes not only the cosmological parameters (the standard set of $w_0 w_a CDM$ cosmological parameters is exposed, using $\sigma_8$ as an amplitude parameter), but also other input quantities such as redshift distributions as described below.

In the rest of this section we describe the cosmological calculations that are implemented using these facilities.
\subsection{Formalism \& Implementation}

\subsubsection{Background cosmology}

The computation of the evolution of the cosmological background follows a typical implementation of a Friedmann equation (\citealp[see e.g.][]{2005A&A...443..819P}):
\begin{equation}
    E^2(a) = \frac{\mathrm{H}^2(a)}{\mathrm{H}^2_0} = \Omega_m a^{-3} + \Omega_k a^{-2} + \Omega_{de} e^{f(a)}
\end{equation}
with $a=1/(1+z)$ the scale factor related to the redshift $z$, $\mathrm{H}(a)=\dot{a}/a$ the Hubble parameter with $\mathrm{H}_0$ its present day value, $\Omega_m= \Omega_{cdm}+\Omega_b$, $\Omega_{de}=1-\Omega_k-\Omega_m$, and 
\begin{equation}
    f(a) = -3 (1 + w_0 + w_a) \ln(a) + 3 w_a (a - 1)
\end{equation}
Notably, the relativistic contributions of massless neutrinos and photon radiation, as well as the massive neutrino contribution are currently neglected. From these expressions, in \texttt{jc.background}, are computed the different cosmological distance functions, such as the radial comoving distance
\begin{equation}
     \chi(a) =  R_H \int_a^1 \frac{\mathrm{d}a^\prime}{{a^\prime}^2 E(a^\prime)}
     \label{eq:radial_comoving}
\end{equation}
with $R_H$ the Hubble radius. 
%Using the scale factor ($a$) redshift ($z$) relationship, $\chi$ can be viewed as a function of $z$. 
%
\subsubsection{Growth of perturbations}
%
Currently, \jaxcosmo\  implements the \citet{Eisenstein_1998} transfer function $T$ which transforms the primordial matter power spectrum to its late-time non-linear value:
\begin{equation}
    P(k, z) = P(k, z=\infty) \cdot T^2(k, z; \Omega_m, \Omega_b, ...),
\end{equation}
through the \textit{halofit} model by \cite{2012ApJ...761..152T} or \cite{2003MNRAS.341.1311S} without the neutrino contribution introduced by \cite{10.1111/j.1365-2966.2011.20222.x}. No Baryon feedback modeling is considered yet.

@@

The normalisation $A$ is parameterised via $\sigma_8$ at $z=0$ as 
\begin{equation}
    A = \sigma_8^2 \times \left(\frac{1}{2 \pi^2} \int_0^\infty \frac{\mathrm{d}k}{k} k^3 P(k) W^2(kR_8) \right)^{-1}
\end{equation}
with $R_8 = 8 \mathrm{Mpc}/h$ and $W(x)$ related to the $j_1$ spherical Bessel function as
\begin{equation}
    W(x) = \frac{3j_1(x)}{x}
\end{equation}

%Future version of the library would offer the possibility to call a \jax\ emulator of the Cosmic Linear Anisotropy Solving System \citep{2011JCAP...07..034B}. \JZ{How close is this to being usable? If it's just an idea or an early prototype then skip this paragraph.}
%
\subsubsection{Angular power spectra}
%
\jaxcosmo\ is currently focused on predicting projected 2D Fourier-space 2pt galaxy lensing, clustering, and cross correlations, the $C_\ell$ angular power spectra that are a primary target of upcoming photometric surveys. The details of the implementation is in \texttt{jc.angular\_cl} which deals with the mean and Gaussian covariance matrix computations.
